Static Analysis by Path Focusing

### 2011-06-13 report/report.tex
minor changes in the report
bug fix in the report
report (final version)

\documentclass[a4paper,english,titlepage,11pt]{report}

\usepackage{verimagstudent}
\usepackage{mathcomp,amssymb}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage[toc,page]{appendix} 
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}

@@

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification is aimed at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that allows to compute an
	approximation of the set of these values, since it is impossible to compute
	the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce rapport s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R internship's report}

@@

Linear Relation Analysis over-approximates the set of possible states as a
convex polyhedron where the dimensions are the numerical variables of the
program.
Instead of convex polyhedra, one could choose intervals, octagons, etc.
The fixpoint equation is computed iteratively; that is, successive
approximations of the set of reachable states are computed until they converge
to a fixpoint.

@@

The fixpoint computation algorithm consists in replacing iteratively each $X_p$ by its value
on the right hand side, until the convergence.

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, $P_w$ can be chosen as the set of the
loop headers.

@@

Acceleration is possible when the loop is simple enough, and comes to compute
the transitive closure $\tau_e^+$ of $\tau_e$, the transition function
associated to the loop.

@@

We apply the linear relation analysis technique over the program in figure
\ref{runningexample}. The set of state where we apply widening is $\{p_1\}$: it
is sufficient, since removing $p_1$ cuts all cycles in the graph.

@@

When computing the fixpoint iterations, we only consider a subset of the graph
for a while, until the convergence. In our example:
\begin{itemize}
\item Step 1:
	we only consider the path of
	the loop that is possible at the first iteration. The second path is
	ignored, and will be treated later. We compute the iterations, and we obtain
	an first invariant for this part of the graph: at point $p_1$, $x =
	y \wedge x \geq 0$.
	Then, we can do some
	narrowing steps in order to recover some precision, before adding new paths
	in the graph. At point $p_1$ after narrowing, $x=y \wedge 0 \leq x \leq 51$.
\item Step 2:
	Now, we add the new paths into our graph if they have become possible. In
	our case, the second path of the loop is now possible, because $x$ can be
	equal to $51$, so we add the path in
	the graph and compute new iterations until convergence\dots. After
	convergence, at point $p_1$, $x + y - 102 \leq 0 \wedge x
	\leq y$. We then apply a narrowing step and at $p_1$, 
$x + y - 102 \leq 0 \wedge x \leq y \wedge y \geq 0$.
\item Step 3: finally, the path from $p_4$ to $p_7$ becomes possible, so we add
it into the graph and compute the invariant for $p_7$.
\end{itemize}

@@

@@

	The multigraph is then the graph such that:
	\begin{itemize}
	\item each element of $P_R$ is a point of the graph.
	\item for each path $p_1 \rightarrow p_2 \rightarrow \dots \rightarrow p_k$
	in the control flow graph, such that $p_1 \in P_R, p_k \in P_R$,
	there is a transition in the multigraph from $p_1$ to $p_k$ with the
	semantic $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
	\rightarrow p_k} \circ
	\dots \circ \tau_{p_1 \rightarrow p_2}$$ where $\tau_{p_i \rightarrow
	p_{i+1}}$ is the transformation associated to the semantic of $p_i
	\rightarrow p_{i+1}$.
	\end{itemize}

@@

  \path [transition] 
		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
		\begin{array}{l}
			x \geq 51 \\
			y \leq 0 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
  \path [transition2] 
		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \geq -1 \\
			y \leftarrow y+1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
		\begin{array}{l}
			x \geq 51 \\
			y \geq 1 \\
			y \leftarrow y-1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \leq -2 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
\end{tikzpicture}
\end{minipage}
\caption{Example of classical control flow graph, and its associated expanded
multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
in the control flow graph corresponds to a simple dashed transition in the multigraph,
having the same semantic.}
\label{multigraphtransformGopan}
\end{figure}
\FloatBarrier

@@

	\subsubsection{Construction of the formula}
	The first point of the Path Focusing technique is to compute the SMT formula
	$\rho$ expressing the semantic of the program. For doing so, all the
	operations in the program have to be expressible within the theory we
	choose. Otherwise, the SMT-solver would not be able to decide if the formula
	is \emph{sat} or not.

@@

\begin{enumerate}
\item Each numerical SSA variable of the control flow graph has its definition
in the $\rho$ formula.
\item For each operation in the program, we encode its semantic, expressed in
terms of the Boolean and numerical SSA variables. For instance, if the CFG
contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ in
the formula. This equality expression is noted $assign(x.2)$. 
$assign(x)$ is defined for each numerical SSA-variable of the program.
\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
$c(i,j)$ expresses the condition that needs to be true to go through the
transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
$c_{i,j}$ is a Boolean variable left non-deterministic.
\item For each point $p_i \notin P_R$, we set the reachability predicate 
$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
\item For each point $p_i^d \in P_R$, we set the reachability predicate 
$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
formula. We simply use $\ite$ (\emph{if-then-else}) statements, a standard feature of the
SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
depending on the incoming transition we come from. Generally, for a
$\Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply apply the widening
	operator,  some narrowing steps could be performed to regain precision
	before focusing on another path. This relates to the ideas in
        \emph{Lookahead widening} \cite{GopanR06}.

	So, the algorithm has a special case if the path we focus on goes from $p_i$
	to itself: instead of computing $\tau_e^\#(X_{p_i})$, one compute
	$SelfLoop(\tau_e^\#,X_{p_i})$, defined in algorithm \ref{loopiteralgo}.

	The $SelfLoop$ procedure actually computes a widening operation, and
	computes a narrowing step in order to recover precision. In some cases,
	instead of performing widening/narrowing iterations, 
	acceleration techniques could be used 
	to find even more precise invariants \cite{GH06}.

@@

	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
	loop, i.e compute the invariants for a loop having its semantic
	expressed by a transition system. The definition of a transition system is
	the following:

	\begin{definition}
	A transition system for a control point $p$ is a DNF formula (i.e a
	disjunction of conjunctions), where each
	disjunct is the semantic of a path from $p$ to itself. It is of the form
	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
	paths, and $\tau_{p,i}$ is the semantic of the $i$-th path from $p$ to
	itself.
	\end{definition}

@@

@@

LLVM internal representation can be applied transformation passes that modifies
parts of it. These passes are mostly used for optimisation purpose. In our
analyser, we apply some already existing passes before running our 
abstract interpretation pass. Among the passes we run, we could cite:
\begin{itemize}
\item the \emph{PromoteMemoryToRegister} Pass, that transforms the internal
representation by promoting the memory variables to Static Single Assignment
(SSA) registers. This pass transform the control flow graph such that all the
variables are in SSA form, and removes most of the \emph{Load/Store} operations
into memory. It is useful since our analyser does not have a memory model, and
then loses lots of precision in the case of \emph{Load/Store} instructions.
\item the \emph{LoopSimplify} and \emph{LoopInfo} Passes, that transform all loops into a canonical
format, and compute some information related to loops (for instance, the set of
BasicBlocks that are parts of the loop, etc.).
\item the \emph{LowerSwitch} Pass, that transform a switch instruction into its
corresponding \emph{if-then-else} instructions. After running this pass, there
is no more \emph{Switch} instruction in the control flow graph, so the
implementation of the static analysis is easier.
\end{itemize}

@@

@@

	\subsection{Dimensions of the Abstract values}

@@

The set of elements $x$ satisfying $x \neq N$ is not convex: it is the union of
two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
represent the set of possible states of our program, we loose information when
the analysis comes across a \emph{diseq} operation.
One classical approach would be to compute a disjunctive invariant, by attaching
two different polyhedron to the BasicBlock. However, we preferred the following
technique: we transform all the operations of the form $x \neq y$ into
$x < y \vee x > y$, meaning that we separate the cases \emph{less than} and
\emph{greater than} into two distinct paths.

@@

\emph{ex6.c} is a program containing non-linear arithmetic: there is an integer
multiplication inside the loop. The Apron library handles multiplication using
a linearization technique. However, multiplications is not allowed in LIA
(Linear Integer Arithmetic) or LRA (Linear Real Arithmetic). Then, Yices
SMT-solver errs with a message ``not implemented'' (which is the reason of the
$\infty$ symbol in the table), when Z3 still works correctly.

@@

	\emph{Path Focusing} technique finds more precise invariants in some cases,
	whereas \emph{Lookahead Widening} performs better in other cases.
	This is partly due to the widening operator: even if the analysis is more
	subtle, the result at the end may be worse.

@@

\begin{itemize}
\item We use the domain of convex polyhedra as the abstract domain. Since this
domain requires a widening operator, a more subtle analysis technique may
sometimes discover invariants that are less precise than a more simplified
technique.
\item Our technique performs well in the case of self loops in the multigraph,
thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
paths that do not constitute a self loop in the multigraph, one performs a
widening operation, but no narrowing iteration. Apparently, lots of the cases
where \emph{Lookahead Widening} performs better than \emph{Path Focusing} are
due to this drawback. An example of such loss of precision is depicted in
appendix \ref{lostprecision}.
An idea that could solve this problem is to try to combine
the two methods (see \ref{combining}).
\end{itemize}

@@

	Validation of software has become an essential domain in industry,
	especially those dealing with embedded systems, where a bug in the program
	can endanger the safety of people, or induce huge costs to fix. These
	domains constitute a privileged application field for formal verification
	techniques.
	Abstract Interpretation is a common used technique to do static analysis of
	programs. It is aimed at computing iteratively an over-approximation of the set of
	possible states of the program, and allows to prove properties such as
	``In this program, $x$ is always between 1 and 10'', ``The program never
	divides by zero'', etc. The objective is to obtain a result as precise as
	possible, by limiting the loss of precision during the analysis, induced for
	instance by the widening operator, or control-flow merges.

@@

	Our analyser is able to deal with small- and middle-sized test-cases, as
	well as real-life programs. It looses
	precision in the case of programs with pointers or several functions.
	It has been tested with program having loops containing several hundreds
	lines of code, and gave results after a time in the order of the second.
	It could be improved in many ways, in order to support
	for instance interprocedural analysis. A memory model could be added to
	handle programs with pointers and \emph{load/store} instructions.
	One could also try to compute disjunctive invariants using the technique
	described in this report.
	Since our technique combines well with acceleration techniques, another
	direction is to implement this last technique to improve our results
	obtained with classical widening/narrowing.

>>>>

\documentclass[a4paper,english,titlepage,11pt,twoside]{report}

\usepackage[pdftex]{graphicx}
\usepackage{mathcomp,amssymb}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath}
\usepackage{mathenv}
\usepackage{tikz}
\usepackage{listings}
\usepackage{placeins}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{multirow}
\usepackage{verimagstudent}
\usepackage[toc,page]{appendix}

\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}

@@

\lstnewenvironment{Csmall}
{\lstset{language=C,
		basicstyle=\ttfamily\small,
		commentstyle=\color{moka}\textit,
		keywordstyle=\color{minuit},
		identifierstyle=\color{trefle},
		showstringspaces=false}}
{}
\title{Static Analysis by Path Focusing}
\author{Julien Henry}
\date{2011}
\shorttitle{Static Analysis by Path Focusing}
\institute{Grenoble-INP}
\abstract{
	Program verification aims at statically discovering properties on programs,
	such as the values that can take the different variables during execution.
	Abstract Interpretation is a technique that  computes an
	over-approximation of the set of these values, since it is impossible to
	compute the real set in general.
	This report takes place in the many attempts to improve the precision of
	static Analysis by Abstract Interpretation. It proposes a technique that
	takes benefit of SMT-solving to obtain more precise results at reasonable
	cost.
}
\resume{  
	La vérification de programme consiste à découvrir statiquement des
	propriétés sur des programmes, comme l'ensemble des valeurs que peuvent
	prendre les variables durant l'exécution. L'Interprétation Abstraite est une
	technique permettant de calculer une approximation de cet ensemble, le
	véritable ensemble étant impossible à calculer en général.
	Ce rapport s'inscrit dans la lignée des travaux visant à améliorer la
	précision de l'analyse par interprétation abstraite, et propose une
	technique tirant parti du SMT-solving pour obtenir de meilleurs résultats
	à un coût raisonnable.
}
\keywords{Static Program Analysis, Verification, Abstract Interpretation,
Invariant Generation, Path Focusing, SMT-Solving}
\motscles{Analyse Statique de Programmes, Vérification, Interprétation
Abstraite, Génération d'invariants, Découverte de Chemins, SMT-Solving}
\tutor{David Monniaux - Matthieu Moy}
%\notes{some notes}
\webaddress{Julien.Henry@imag.fr}
\reporttype{M2R Internship Report}

@@

Linear Relation Analysis over-approximates the set of possible states as a
convex polyhedron where the dimensions are the numerical variables of the
program.
Instead of convex polyhedra, one could choose intervals, octagons, etc, that are
subclasses of the class of convex polyhedra.
The fixpoint equation is computed iteratively; that is, successive
approximations of the set of reachable states are computed until they converge
to a fixpoint.

@@

The fixpoint computation algorithm consists in replacing iteratively each $X_p$ by its value
on the right hand side, until convergence.

@@

\subsection{Precision of the analysis}
In order to guarantee the termination of the computation, there is no need to
apply the widening operator at each control point $p \in P$. It is sufficient to
apply it on a subset of $P$, called $P_w$, such that removing the nodes of $P_w$
disconnects every cycles. For instance, in structured programs, $P_w$ can be chosen as the set of the
loop headers.

@@

Acceleration is possible when the loop is simple enough, and comes to computing
the transitive closure $\tau_e^+$ of $\tau_e$, the transition function
associated to the loop.

@@

We apply the linear relation analysis technique over the program in figure
\ref{runningexample}. The set of control state where we apply widening is $\{p_1\}$: it
is sufficient, since removing $p_1$ cuts all cycles in the graph.

@@

When computing the fixpoint iterations, we only consider a subset of the graph
for a while, until the convergence. In our example:
\begin{itemize}
\item Step 1:
	we only consider the path of
	the loop that is possible at the first iteration. The second path is
	ignored, and will be treated later. We compute the iterations, and we obtain
	an first invariant for this part of the graph: at point $p_1$, $x =
	y \wedge x \geq 0$.
	Then, we can do some
	narrowing steps in order to recover some precision, before adding new paths
	in the graph. At point $p_1$ after narrowing, $x=y \wedge 0 \leq x \leq 51$.
\item Step 2:
	Now, we add the new paths into our graph if they have become possible. In
	our case, the second path of the loop is now possible, because $x$ can be
	equal to $51$, so we add the path to
	the graph and compute new iterations until convergence\dots. After
	convergence, at point $p_1$, $x + y - 102 \leq 0 \wedge x
	\leq y$. We then apply a narrowing step and at $p_1$, 
$x + y - 102 \leq 0 \wedge x \leq y \wedge y \geq 0$.
\item Step 3: finally, the path from $p_4$ to $p_7$ becomes possible, so we add
it into the graph and compute the invariant for $p_7$.
\end{itemize}

@@

	In this section, we present a refinement of Abstract Interpretation, refered
	as \emph{Path Focusing} technique \cite{MG11}.

@@

	The multigraph is then the graph such that:
	\begin{itemize}
	\item each element of $P_R$ is a point of the graph.
	\item for each path $p_1 \rightarrow p_2 \rightarrow \dots \rightarrow p_k$
	in the control flow graph, such that $p_1 \in P_R, p_k \in P_R$,
	there is a transition in the multigraph from $p_1$ to $p_k$ with the
	semantics $$\tau_{p_1 \rightarrow \dots \rightarrow p_k} = \tau_{p_{k-1}
	\rightarrow p_k} \circ
	\dots \circ \tau_{p_1 \rightarrow p_2}$$ where $\tau_{p_i \rightarrow
	p_{i+1}}$ is the transformation associated to the semantics of $p_i
	\rightarrow p_{i+1}$.
	\end{itemize}

@@

  \path [transition] 
		(n0) edge              node [yshift=5mm] {$x,y \leftarrow 0$} (n1);
  \path [transition] 
		(n1) edge    [bend left]  node [right,yshift=-0.8cm]  {$
		\begin{array}{l}
			x \geq 51 \\
			y \leq 0 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
  \path [transition2] 
		(n1) edge    [loop left,distance=1.2cm]  node [left,xshift=3mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \geq -1 \\
			y \leftarrow y+1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [loop right,distance=1.2cm]  node [right,xshift=-2mm] {$
		\begin{array}{l}
			x \geq 51 \\
			y \geq 1 \\
			y \leftarrow y-1 \\
			x \leftarrow x+1 
		\end{array}
		$} (n1);
  \path [transition] 
		(n1) edge    [bend right]  node [left,yshift=-0.8cm,xshift=4mm] {$
		\begin{array}{l}
			x \leq 50 \\
			y \leq -2 \\
			y \leftarrow y-1
		\end{array}
		$} (n7);
\end{tikzpicture}
\end{minipage}
\caption{Example of classical control flow graph, and its associated expanded
multigraph, where $P_R = \{p_0, p_1, p_7\}$. The path dashed from $p_1$ to $p_1$
in the control flow graph corresponds to a simple dashed transition in the multigraph,
having the same semantics.}
\label{multigraphtransformGopan}
\end{figure}
\FloatBarrier

@@

	\subsubsection{Construction of the formula}
	The first point of the Path Focusing technique is to compute the SMT formula
	$\rho$ expressing the semantics of the program. For doing so, all the
	operations in the program have to be expressible within the theory we
	choose. Otherwise, the SMT-solver would not be able to decide if the formula
	is \emph{sat} or not.

@@

\begin{enumerate}
\item Each numerical SSA variable of the control flow graph has its definition
in the $\rho$ formula.
\item For each operation in the program, we encode its semantics, expressed in
terms of the Boolean and numerical SSA variables. For instance, if the CFG
contains the assignment $x.2 = x.1 + 1$, then we simply add $x.2 = x.1 + 1$ to
the formula. This equality expression is noted $assign(x.2)$. 
$assign(x)$ is defined for each numerical SSA-variable of the program.
\item For each transition $(p_i,p_j) \in E$, we set $t_{i,j} = B_i\ \wedge \
c(i,j)$, where $B_i = b_i^s$ if $p_i \in P_R$, or $b_i$ if not.
$c(i,j)$ expresses the condition that needs to be true to go through the
transition. For instance, $c(i,j)$ could be of the form $x < y$, \dots
If the transition is non-deterministic, then $c(i,j) = c_{i,j}$, where
$c_{i,j}$ is a Boolean variable left non-deterministic.
\item For each point $p_i \notin P_R$, we set the reachability predicate 
$b_i = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$.
\item For each point $p_i^d \in P_R$, we set the reachability predicate 
$b_i^d = \displaystyle \bigvee_{(p_j,p_i)\in E} t_{j,i}$. 
\item Numerical $\Phi$-variables assignations have to be expressed in the $\rho$
formula. We simply use $\ite$ (\emph{if-then-else}) statements, a standard feature of the
SMT-Lib format \cite{BarST-SMTLIB}, for giving the right value to the variable,
depending on the incoming transition we come from. Generally, for a
$\Phi$-variable $x = \Phi(x_1,x_2,\dots,x_n)$ in control point $p_i$ with the
ordered incoming transitions $t_{j_1,i},t_{j_2,i},\dots,t_{j_n,i}$, we set:

@@

	One can refine the \emph{Path-focusing} algorithm by treating specially the
	self loops of the multigraph. Indeed, instead of simply applying the widening
	operator,  some narrowing steps could be performed to regain precision
	before focusing on another path. This relates to the ideas in
        \emph{Lookahead widening} \cite{GopanR06}.

	So, the algorithm has a special case if the path we focus on goes from $p_i$
	to itself: instead of computing $\tau_e^\#(X_{p_i})$, one compute
	\textit{SelfLoop}$(\tau_e^\#,X_{p_i})$, defined in algorithm \ref{loopiteralgo}.

	The \textit{SelfLoop} procedure actually computes a widening operation, and
	computes a narrowing step in order to recover precision. In some cases,
	instead of performing widening/narrowing iterations, 
	acceleration techniques could be used 
	to find even more precise invariants \cite{GH06}.

@@

	\cite{GulwaniZ10} proposes a technique to compute transitive closure of a
	loop, i.e compute the invariants for a loop having its semantics
	expressed by a transition system. The definition of a transition system is
	the following:

	\begin{definition}
	A transition system for a control point $p$ is a DNF formula (i.e a
	disjunction of conjunctions), where each
	disjunct is the semantics of a path from $p$ to itself. It is of the form
	$$\bigvee_{1 \leq i \leq n} \tau_{p,i}$$ where $n$ is the number of
	paths, and $\tau_{p,i}$ is the semantics of the $i$-th path from $p$ to
	itself.
	\end{definition}

@@

Computing disjunctive invariants is a field we have to work on. The work
described here is only preliminary, and shows it is possible to improve existing
techniques with our ideas, so that we avoid explicit enumeration of
exponential-sized sets of paths.

@@

LLVM internal representation can be applied transformation passes that modifies
parts of it. These passes are mostly used for optimisation purpose. In our
analyser, we apply some already existing passes before running our 
abstract interpretation pass. Among the passes we run, we could cite:
\begin{itemize}
\item the \emph{PromoteMemoryToRegister} Pass (-mem2reg), that transforms the internal
representation by promoting the memory variables to Static Single Assignment
(SSA) registers. This pass transform the control flow graph such that all the
variables are in SSA form, and removes most of the \emph{Load/Store} operations
into memory. It is useful since our analyser does not have a memory model, and
then loses lots of precision in the case of \emph{Load/Store} instructions.
\item the \emph{LoopSimplify} (-loopsimplify) and \emph{LoopInfo} (-loopinfo) Passes, that transform all loops into a canonical
format, and compute some information related to loops (for instance, the set of
BasicBlocks that are parts of the loop, etc.).
\item the \emph{LowerSwitch} Pass (-lowerswitch), that transform a switch instruction into its
corresponding \emph{if-then-else} instructions. After running this pass, there
is no more \emph{Switch} instruction in the control flow graph, so the
implementation of the static analysis is easier.
\end{itemize}

@@

	Morally, this means the $\Phi$-variables are on incoming transitions,
	whereas the other instructions are on outgoing transitions.

@@

	\subsection{Dimensions of the abstract values}

@@

The set of elements $x$ satisfying $x \neq N$ is not convex: it is the union of
two intervals: $]-\infty, N[\  \cup\  ]N,+\infty[$. Since we use convex polyhedra to
represent the set of possible states of our program, we lose information when
the analysis comes across a \emph{diseq} operation.
One classical approach would be to compute a disjunctive invariant, by attaching
two different polyhedron to the BasicBlock. However, we preferred the following
technique: we transform all the operations of the form $x \neq y$ into
$x < y \vee x > y$, meaning that we separate the cases \emph{less than} and
\emph{greater than} into two distinct paths.

@@

\emph{ex6.c} is a program containing non-linear arithmetic: there is an integer
multiplication inside the loop. The Apron library handles multiplication using
a linearization technique. However, multiplication is not allowed in LIA
(Linear Integer Arithmetic) or LRA (Linear Real Arithmetic). Then, Yices
SMT-solver errs with a message ``not implemented'' (which is the reason of the
$\infty$ symbol in the table), when Z3 still works correctly.

@@

	\emph{Path Focusing} technique finds more precise invariants in some cases,
	whereas \emph{Lookahead Widening} performs better in other cases.
	This is partly due to the widening operator: even if the analysis is more
	subtle, the result at the end may be worse. This is a well-known result in
	abstract interpretation: when using widening operations, the final result is
	not monotonic with the quality of the abstraction.

@@

\begin{itemize}
\item We use the domain of convex polyhedra as the abstract domain. Since this
domain requires a widening operator, a more subtle analysis technique may
sometimes discover invariants that are less precise than a more simplified
technique.
\item Our technique performs well in the case of self loops in the multigraph,
thanks to the widening/narrowing phases (or acceleration). Yet, in the case of
paths that do not constitute a self loop in the multigraph, one performs a
widening operation, but no narrowing iteration. Apparently, lots of the cases
where \emph{Lookahead Widening} performs better than \emph{Path Focusing} are
due to this drawback. An example of such loss of precision is depicted in
appendix \ref{lostprecision}.
An idea that could solve this problem is to try to combine
the two methods, and get the best of both worlds (see \ref{combining}).
\end{itemize}

@@

	Validation of software has become an essential domain in industry,
	especially those dealing with embedded systems, where a bug in the program
	can endanger the safety of people, or induce huge costs to fix. These
	domains constitute a privileged application field for formal verification
	techniques.
	Abstract Interpretation is a commonly used technique to do static analysis of
	programs. It is aimed at computing iteratively an over-approximation of the set of
	possible states of the program, and allows to prove properties such as
	``In this program, $x$ is always between 1 and 10'', ``The program never
	divides by zero'', etc. The objective is to obtain a result as precise as
	possible, by limiting the loss of precision during the analysis, induced for
	instance by the widening operator, or control-flow merges.

@@

	Our analyser is able to deal with small- and middle-sized test-cases, as
	well as real-life programs. It loses
	precision in the case of programs with pointers or several functions.
	It has been tested with program having loops containing several hundreds
	lines of code, and gave results after a time in the order of the second.
	It could be improved in many ways, in order to support
	for instance interprocedural analysis. A memory model could be added to
	handle programs with pointers and \emph{load/store} instructions.
	One could also try to compute disjunctive invariants using the technique
	described in this report.
	Since our technique combines well with acceleration techniques, another
	direction is to implement this last technique to improve our results
	obtained with classical widening/narrowing.
